Red Wine Quality by Andrew Phan
========================================================
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

getwd()
setwd("/Users/andrewphan/Downloads")

library(gridExtra)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
```

```{r echo=FALSE, Load_the_Data}

# Load the Data
rw = read.csv('wineQualityReds.csv')

```

I am going to be exploring a data set about 1599 red wines and 11 of 
of their chemical properties. To go along with the properties, overall 
quality ratings, which were medians of ratings from 0-10 from at least three 
wine experts, were included for each red wine. 

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots: Histograms}

ggplot(aes(x=alcohol), data=rw)+
  geom_histogram(binwidth=0.5)+
  scale_x_continuous(breaks=seq(0,15,0.5))

summary(rw$alcohol)

```

The lowest alcohol percentage a red wine had was 8.4%, and the highest was 
14.9%. It seems that the distribution for alcohol percentage, by volume, 
is skewed right. There seems to be a very high count of red wines which are
approximately 9.5% alcohol. 

```{r echo=FALSE}

ggplot(aes(x=pH), data=rw)+
  geom_histogram() +
  scale_x_continuous(breaks=seq(0,4,0.05))

summary(rw$pH)

```

The pH levels for the red wines seem to follow a normal distribution pattern.
It seems there are a few possible outliers around 3.85 pH and 2.75 pH. Nothing
too out of the ordinary.

```{r echo=FALSE}

ggplot(aes(x=quality), data=rw)+
  geom_histogram(binwidth=1)+
  scale_x_continuous(breaks=seq(0,10,1))

```

I wanted to take a look at the distribution of quality ratings for the red
wines to make sure nothing was out of the ordinary. This histogram gives us
a good idea of whether the experts were being too harsh, or too easy. The 
normal distribution for quality ratings with no ratings below 3 and none 
above 8 gives me the reassurance that the experts were not being overly 
critical or easygoing.

```{r echo=FALSE, Univariate_Plots: Boxplots}

ggplot(aes(x="", y=pH), data=rw)+
  geom_boxplot()

```

I made a boxplot for pH so I could see if the outliers were having a 
significant effect on the distribution of the data; they don't seem to be much
of a problem based on the boxplot. 


```{r echo = FALSE}

ggplot(aes(x="", y=fixed.acidity), data=rw)+
  geom_boxplot()

summary(rw$fixed.acidity)

```

The boxplot for fixed acidity seems to show there are a lot of outliers above
1.5 * IQR. It might be a good idea to see what it looks like with the outliers
removed.

Most red wines have fixed acidity levels between 7.10 and 9.20. The IQR is 
2.1. 1.5*IQR = 3.15. Removing the cases where the fixed acidity levels are 
above Q3+3.15 (outliers) might give us a better look at the distribution. 


```{r echo = FALSE}

sub_acid = subset(rw, rw$fixed.acidity < 9.20+3.15)
# Subsets the data, so that only red wines with fixed acidity levels below
# 9.20 + 3.15 are considered

ggplot(aes(x="", y=fixed.acidity), data=sub_acid)+
  geom_boxplot()

```

The distribution seems to look more normal now. With the knowledge that there 
are only outliers above 1.5*IQR, perhaps we can take a look at the histogram. 
I would expect the distribution to look slightly skewed right, as it seems the
there are a good number of outliers above Q3 + 1.5*IQR.


```{r echo=FALSE}

ggplot(aes(x=fixed.acidity), data=rw)+
  geom_histogram()

```

The prediction was correct!

# Univariate Analysis

### Structure of my dataset:

The dataset has 13 columns, and 1599 rows. Something to note is one of the
columns, X, is just an indexing column. It can likely be deleted. I couldn't
find anything in the documentation about the columns about the X column. It is 
likely the column is redundant.

### Main feature of interest:

The main feature of interest in this dataset is the quality column. I'm 
interested in which of the measured chemical properties have the most influence
on the quality rating. 

Aside from that, I'm interested in the relationships between some of the
chemical properties. 

### Extra notes about the dataset:

The dataset is fairly clean and tidy. Each variable is a column, and each
observation is a row. There doesn't seem to be any data quality issues, either.

### Unusual distributions/findings: 

As noted in one of my comments, fixed acidity levels had a skewed right
distribution. Alcohol percentage by volume is also skewed right, with an 
extremely high count of red wines with approximately 9.5% percentage alcohol.





# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots: Quality}

ggplot(aes(x=quality,y=alcohol), data=rw)+
  geom_jitter(alpha=0.25)+
  scale_y_continuous(breaks=seq(0,16,1))+
  scale_x_continuous(breaks=seq(3,8,1))
  
cor.test(rw$alcohol,rw$quality)

```

It seems there is a moderate positive relationship between alcohol percentage
and quality. I wanted to dig deeper and examine if the higher quality 
wines were focused around certain alcohol percentage levels. 

```{r echo = FALSE}

ggplot(aes(x=alcohol), data=subset(rw,rw$quality==7)) +
  geom_histogram()

```

The distribution of alcohol percentage levels for red wines which
were rated 7 is normal. 

```{r echo = FALSE}

ggplot(aes(x=alcohol),data = subset(rw,rw$quality ==8))+ 
  geom_histogram(binwidth=0.6)

```

The distribution, for the most part, looks normal based on the small number 
of observations we have.

```{r echo=FALSE}

ggplot(aes(x=quality, y =fixed.acidity),data=rw)+
  geom_jitter(alpha=0.25)+
  scale_x_continuous(breaks=seq(3,8,1))

cor.test(rw$fixed.acidity,rw$quality)

```

There doesn't seem to any type of significant relationship between fixed 
acidity levels and quality based on the correlation coefficient of 0.124.
The fixed acidity levels for red wines rated an 8 seem to be more focused 
within a certain range, around 7-10.5 g/dm^3. This might not be very telling,
since again, there aren't too many red wines with a rating of 8. I'm going
to compare the boxplots for fixed acidity for wines of different quality 
ratings.

```{r echo = FALSE}

p1 = ggplot(aes(x="7",y=fixed.acidity),data = subset(rw,rw$quality ==7))+ 
  geom_boxplot()+
  scale_y_continuous(breaks=seq(0,20,1))+
  coord_cartesian(ylim=c(0,20))

p2 = ggplot(aes(x="8",y=fixed.acidity),data = subset(rw,rw$quality ==8))+ 
  geom_boxplot()+
  scale_y_continuous(breaks=seq(0,20,1))+
  coord_cartesian(ylim=c(0,20))

p3 = ggplot(aes(x="3",y=fixed.acidity),data = subset(rw,rw$quality ==3))+ 
  geom_boxplot()+
  scale_y_continuous(breaks=seq(0,20,1))+
  coord_cartesian(ylim=c(0,20))

p4 = ggplot(aes(x="4",y=fixed.acidity),data = subset(rw,rw$quality ==4))+ 
  geom_boxplot()+
  scale_y_continuous(breaks=seq(0,20,1))+
  coord_cartesian(ylim=c(0,20))

```

It seems that the wines of quality 7 had fixed acidity levels from
around 7-10.5 g/dm^3. It might be a good idea to now examine the boxplots of
lower quality wines next to these.Though 75% of the wines with rating 3 seemed 
to have acidity levels between 7.5-10 g/dm^3, it should be noted that the lower 
whisker for these wines is very short. The wines with rating 4 actually seemed 
to be more packed into a certain range: 7-8.5 g/dm^3.

```{r echo=FALSE}

ggplot(aes(x=quality, y =residual.sugar),data=rw)+
  geom_jitter(alpha=0.25)+
  scale_x_continuous(breaks=seq(3,8,1))

cor.test(rw$residual.sugar,rw$quality)

```

Based on this scatterplot, it seems to me that wines with higher quality
ratings did not have residual sugar levels above 9. This could be a sign that
the fermentation process which results in the best quality red wines
also keeps the residual sugar levels fairly and consistently low. However, 
it doesn't look like there are too many wines with residual sugar levels 
above 9, even for red wines with quality ratings lower than 7. Interestingly
enough, the wines with rating 3 also had no residual sugar levels above 9.
This may just be a result of not enough observations for wines of those 
quality ratings. The correlation coefficient does not suggest a significant
linear relationship exists between residual sugar levels and quality ratings.


```{r echo = FALSE}

ggplot(aes(x=quality, y =sulphates),data=rw)+
  geom_jitter(alpha=.25)+
  scale_y_continuous(breaks=seq(0,2,0.25))+
  scale_x_continuous(breaks=seq(3,8,1))

cor.test(rw$sulphates,rw$quality)

```

Again, the wines of highest quality ratings, 7 and 8, seem to have more
consistency with their sulphate levels. Though it doesn't seem like 
we have found any one property that will give us a good idea of the quality
ratings, it might be true that higher quality wines tend to have more 
consistency. This makes sense, as you would expect the process of creating
the highest quality wines to be very meticulous and careful: a craft of sorts.
It seems reasonable to assume that the entire process for creating the higher
quality wines leads to more consistency. Perhaps this is what I should be 
looking for in the next few plots. Being less careful with the process likely
leads to lower quality wines and less consistency. It should be noted, 
however, that wines of fairly low quality with ratings 3 and 4 also seemed to 
have more consistency than wines with ratings 5 or 6. This, again, points 
to there actually just not being enough wines with ratings 3, 4, 7, or 8.
The correlation coefficient is not significant.

```{r echo=FALSE}

ggplot(aes(x=quality, y =density),data=rw)+
  geom_jitter(alpha=.25)+
  scale_y_continuous(breaks=seq(0,1.01,0.00125))+
  scale_x_continuous(breaks=seq(3,8,1))

cor.test(rw$density,rw$quality)

```

It seems the lower quality wines, wines of rating 3, 4, and 5 tended to stay
above certain density levels. They never strayed below 0.99250 g/cm^3. 
Wines of ratings 6, 7, and 8, however, all had instances where the density 
strayed below 0.99250. Again, the correlation coefficient did not point to
a significant relationship between the variables.

```{r echo=FALSE}

ggplot(aes(x=quality, y = chlorides),data=rw)+
  geom_jitter(alpha=.25)+
  scale_x_continuous(breaks=seq(3,8,1))

cor.test(rw$chlorides,rw$quality)

```

Here, the red wines of higher quality again seemed to be more consistent with
their chloride levels (especially those with a rating of 8). Also, it does 
LOOK LIKE higher quality wines tended to have lower chloride levels.

However, the correlation coefficient does not indicate a significant 
relationship.

It shoud be noted, again, that our findings about consistency might be a 
result of there simply being more wines with ratings 5 and 6. 
We can do a quick check if this is the case with the table command.


```{r echo=FALSE}

table(rw$quality)

```

Indeed, it seems that just was a lot more wines with ratings 5 or 6 versus 
wines rated 3, 4, 7, or 8, which is why it appeared that there was more 
consistency among the more highly rated wines.

```{r echo=FALSE, Bivariate_Plots: Properties}

ggplot(aes(x=free.sulfur.dioxide,y=total.sulfur.dioxide),data=rw)+
  geom_point(alpha=0.2)

cor.test(rw$free.sulfur.dioxide,rw$total.sulfur.dioxide) 

```

Not a very interesting relationship, but it does seem like the higher free
sulfure dioxide levels is positively correlated with higher total sulfur 
dioxide levels. However, it does seem that a great proportion of the red 
wines have fairly low free and total sulfur dioxide levels, indicated by the 
dark blobs still present towards the bottom left of the plot.

It might be interesting to check for relationships between density and 
variables that might contribute to higher density, like presence of sulphates,
residual sugars, chlorides, etc.

```{r echo=FALSE}

ggplot(aes(x=sulphates,y=density),data=rw)+
  geom_point(alpha=0.5)

cor.test(rw$sulphates,rw$density)

```

It does not seem like the presence of higher sulphate levels contributed to 
higher density. 

```{r echo=FALSE}

ggplot(aes(x=residual.sugar,y=density),data=rw)+
  geom_point(alpha=0.5)

cor.test(rw$residual.sugar,rw$density)

```

Based only on the plot, it didn't seem to me like the presence of higher 
residual sugar levels had much of an effect on density. 
Actually, it turns out there is a moderate positive relationship between 
residual sugar levels and density, based on the returned correlation 
coefficient: 0.355. 

```{r echo=FALSE}

ggplot(aes(x=alcohol,y=density),data=rw)+
  geom_point(alpha=0.5)

cor.test(rw$alcohol,rw$density)

```

Interestingly, there does seem to be a relationship between alcohol percentage
levels and density. It seems that as alcohol percentage levels increase, 
density decreases: these variables are negatively correlated. It might be a 
good idea to ask why this makes sense in the analysis.


# Bivariate Analysis

### Relationships observed with main feature of interest:

My main feature of interest was quality. It didn't really seem to me that any 
one chemical property other than alcohol could be said to be associated with 
higher/lower quality ratings. However, something I did notice was that wines 
with higher ratings tended to have more consistency with their measured 
properties than wines ofquality 5 or 6. To reiterate, I thought this would make 
sense because you would expect the winemaking process for the highest 
quality wines to be very precise and  painstaking. This would leave little room 
for inconsistency in the process, and consequently, little room for much 
inconsistency with the chemical properties. Unfortunately, this is probably just
because there were not as many wines with ratings 7 or 8 versus wines rated
5 or 6.

### Relationships observed between other features:

I observed a positive relationship between free sulfur dioxide and total sulfur
dioxide levels.

Though I didn't immediately notice from the plot, there was a moderate positive
relationship between residual sugar level and density. This makes sense;
I expect higher amounts of leftover sugar (denser than water) to contribute
to higher density levels.

Also, there was a negative correlation between density and alcohol
levels. After a quick google search, I found that alcohol is less dense than 
water. This may serve to explain why wines with higher alcohol percentage levels
tended to have lower density.

### Strongest relationship:

The strongest relationship I found was between free sulfur dioxide and total 
sulfur dioxide. The correlation coefficient returned was 0.667.



# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}

sub <- subset(rw, quality >= 4 & quality < 8) 

ggplot(aes(x=volatile.acidity,y=residual.sugar,color=as.factor(quality)),
       data=sub)+ 
    geom_jitter(alpha=0.5) +
    scale_y_continuous(limits = c(quantile(rw$residual.sugar, 0.05), 
                                  quantile(rw$residual.sugar, 0.90)))+
    scale_x_continuous(limits = c(quantile(rw$volatile.acidity, 0.05), 
                                  quantile(rw$volatile.acidity, 0.90)))+
  scale_color_brewer(type='div',palette=7,guide=
                       guide_legend(title='Quality Levels'))

# Subsetted the red wines to only examine wines of quality ratings 4-7.
# Used quality as color, but converted to a factor first.
# Removed lower 5% and upper 10% of residual sugar levels and volatile acidity
# levels. Used scale_color_brewer with diverging palette so it is easier to
# distinguish between the points for the different ratings.

```

It looks like red wines of quality ratings 5 and 6 were more spread out
in terms of volatile acidity and residual sugar levels, while red wines
of quality rating 7 seemed to be focused around the lower volatile acidity
levels. Again, it seems these are the results of there being more red wines
rated 5 or 6 than 7.

```{r echo=FALSE}

ggplot(aes(x=quality,y=alcohol, color=density),data=rw)+
  geom_jitter(alpha=.75)+
  scale_x_continuous(breaks=seq(3,8,1))

ggplot(aes(x=quality,y=residual.sugar, color=density),data=rw)+
  geom_jitter(alpha=.75)+
  scale_x_continuous(breaks=seq(3,8,1))

ggplot(aes(x=quality,y=chlorides, color=sulphates),data=rw)+
  geom_jitter(alpha=.75)+
  scale_x_continuous(breaks=seq(3,8,1))

```

I didn't really find the above three plots very educational. I don't notice
any clear trends or interesting relationships which weren't already 
discussed in my bivariate plot discussion.

```{r echo=FALSE}

ggplot(aes(x=alcohol,y=residual.sugar,color = density),data=rw)+
  geom_point()

```

Based on this scatterplot, it seems that higher alcohol levels along with 
low residual sugar levels were associated with lower density levels. Again,
this makes sense based on the bivariate plots of alcohol vs. density and 
residual sugar levels vs. density.


# Multivariate Analysis

### Relationships observed:

There seemed to be more consistently low volatile acidity levels for red wines
of quality rating 7 compared to red wines rated 5 or 6. Again, this is likely
a result of there simply being many more 5's and 6's versus 7's.

The relationship between residual sugar levels, alcohol, and density were
made even more clear with the multivariate plot of the three variables. Lower
alcohol levels tended to be associated with higher density. Higher alcohol 
levels tended to be associated with lower density. The opposite held for
residual sugar levels.

### Interesting/Surprising Interactions:

The most interesting interaction between features was between residual sugar,
alcohol, and density. We did already know that there was a positive 
relationship between residual sugar levels and density, and a negative
relationship between alcohol and density. We examined the
multivariate scatter plot of alcohol vs. residual sugar levels, with the colors
of the points varying with density. The plot reinforced our findings from our
bivariate plots; the red wines with the higher alcohol levels and lower
residual sugar levels tended to have lower densities, and vice versa.



# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

ggplot(aes(x=quality), data=rw)+
  geom_histogram(binwidth=1,fill='orange',color='black')+
  scale_x_continuous(breaks=seq(0,10,1))+
  xlab('Quality Rating')+
  ylab('Count')+
  ggtitle('Quality Ratings Histogram')

```

### Description One
I believed one of the most important distributions to examine was that of the 
quality ratings. I found it fairly significant that the distribution was normal,
as this signaled the wine experts were not being overly critical or easygoing
when rating the wines. 

### Plot Two
```{r echo=FALSE, Plot_Two}

ggplot(aes(x=alcohol,y=density),data=rw)+
  geom_point(alpha=0.5,colour='brown')+
  scale_y_continuous(breaks=seq(0,1.1,0.00175))+
  xlab('Alcohol Content by Volume (%)')+ 
  ylab('Density (g/cm^3)')+
  ggtitle('Alcohol vs. Density Scatterplot')

```

### Description Two
I chose this plot because it illustrated a relationship between chemical 
properties which could be easily explained and understood. As alcohol percentage 
level increased, density tended to decrease. Again, this is because alcohol is 
not very dense (less dense than water). Consequently, the higher the alcohol 
content, the less dense the red wine, and vice versa.

### Plot Three
```{r echo=FALSE, Plot_Three}

ggplot(aes(x=alcohol,y=residual.sugar,color= density),data=rw)+
  geom_point()+
  scale_y_continuous(breaks=seq(0,16,2))+
  scale_x_continuous(breaks=seq(0,18,1.5))+
  xlab('Alcohol Content by Volume (%)')+
  ylab('Residual Sugar (g/dm^3)')+
  labs(color='Density (g/cm^3)')+
  ggtitle('Residual Sugar vs. Alcohol Content vs. Density')+
  scale_color_gradient(low='grey23',high='orangered')

```

### Description Three
I chose this plot because it brought together my findings from two of my 
bivariate plots. As stated above, I found a negative correlation between 
alcohol content and density. In the bivariate plots section, I found a 
positive relationship between residual sugar levels and density. It then made
sense to create a multivariate plot of alochol content vs. residual sugar levels
with density as color, as it would demonstrate both relationships at once in
one plot. Checking the plot, you can see that red wines with 
high alcohol content and low residual sugar levels tended to have lower
densities, and vice versa. Both relationships are demonstrated in one nice plot.

------

# Reflection

Overall, the code required for this project was not too difficult. I would say
I had a lot of difficulties in a few areas. First, I had a hard time "exploring"
different plots; if I thought that comparing two variables didn't really make 
sense then I would have difficulty justifying creating the plots. This made it
hard to find interactions/relationships that I didn't already know about. 
Second, I didn't really feel like I found many meaningful relationships for
my most important variable: quality. Of course, that could just mean that 
it is a combination of all the properties that makes a wine "good". However, it
was still discouraging to find little to no very interesting plots which 
measured quality as one of the variables, other than alcohol vs. quality.
I also had a lot of trouble figuring out exactly which variables to plot for 
the multivariate plots, and why. Lastly, I am still having some trouble 
actually analyzing the meaning of the plots after I generate them. 
This might just have to do with the relationships I chose to examine being 
insignificant. Something that was surprising to me was how easy it was to clean 
up the plots with labels and colors/outlines. In the future, it might be 
beneficial to create a model to predict if a red wine will be of high 
quality based on given chemical properties.


